\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{geometry}
\geometry{letterpaper, margin=1in}

\usepackage{times}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{appendix}

\urlstyle{same}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
    pdftitle={MinEnglish Language Specification v2.0},
    pdfauthor={Dan Micsa, PhD},
}

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!10},
    keywordstyle=\bfseries,
    commentstyle=\itshape
}

\newcommand{\me}[1]{\texttt{#1}}
\newcommand{\exrow}[4]{%
  \hline
  \textit{English} & #1 & #2 \\
  \textit{MinEnglish} & #3 & #4 \\
}

\title{\Huge\textbf{MinEnglish Language Specification (v2.0)}\\[0.3em]
\LARGE A Zero-Redundancy Phonetic Communication Protocol}
\author{\textbf{Dan Micsa, PhD}\\
\small MIT License (see Appendix)}
\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
MinEnglish is a formal communication protocol engineered to maximize Shannon information density while minimizing cognitive load and parser ambiguity. By systematically eliminating morphological irregularities and orthographic redundancies inherent in natural languages, MinEnglish achieves consistent computational predictability. This specification defines the phonological, morphological, and syntactic architecture of version 2.0, documents a 35-sentence comparative corpus demonstrating an average 35.0\% character reduction over standard American English, and addresses the primary theoretical limitation of the system: biological adoption friction.
\end{abstract}

%---
\section{Design Principles and Underlying Rationale}

To optimize human-to-human and human-to-machine transmission, MinEnglish operates on five absolute constraints:

\begin{itemize}[leftmargin=*]
    \item \textbf{Phonetic Isomorphism:} 1-to-1 grapheme-phoneme correspondence eliminates the cognitive overhead of orthographic memorization and provides predictable tokenization for machine parsers.
    \item \textbf{Morphological Parsimony:} Pluralization, verb conjugation, and declension are replaced by explicit numerical and temporal prefixes, removing ambiguity in morphological boundaries.
    \item \textbf{ASCII Strictness:} Operates exclusively within the standard US keyboard namespace without diacritics. The letters \textit{q}, \textit{w}, and \textit{y} are omitted to balance acoustic distinctness with key space economy.
    \item \textbf{Operator Integration:} Mathematical symbols (\me{*}, \me{\~{}}, \me{>}, \me{<}, \me{:}) function as first-class syntactic tokens, compressing complex semantic modifiers into single-byte representations.
    \item \textbf{Absolute Regularity:} A zero-exception framework guarantees that programmatic rules apply universally, ensuring high predictability for both human acquisition and natural language processing systems.
\end{itemize}

%---
\section{Phonology and Orthography}

\subsection{Vowels}
The vowel space is constrained to five cardinal points to ensure acoustic discernibility.

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{llXl}
\toprule
\textbf{Graph.} & \textbf{Sound} & \textbf{Example} & \textbf{MinEnglish} \\
\midrule
\textbf{a} & as in c\textit{a}t & cat & cat \\
\textbf{e} & as in b\textit{e}d & bed & bed \\
\textbf{i} & as in b\textit{i}t & bit & bit \\
\textbf{o} & as in h\textit{o}t & hot & hot \\
\textbf{u} & as in b\textit{u}t & but & but \\
\bottomrule
\end{tabularx}
\caption{Cardinal Vowel Mapping}
\end{table}

\subsection{Long Vowels and Diphthongs}
Diphthongs are encoded as contiguous grapheme pairs, preserving the 1-to-1 spelling mapping without macron diacritics: \me{ei} (day), \me{ii} (see), \me{ai} (like), \me{ou} (go), \me{uu} (food), \me{au} (how), \me{oi} (boy).

\subsection{Consonants}

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{llX}
\toprule
\textbf{Graph.} & \textbf{Sound} & \textbf{Application Rule} \\
\midrule
\textbf{c}  & /k/ unaspirated & Soft velar plosive: cat, cum, bac \\
\textbf{k}  & /k/ aspirated   & Hard velar plosive: king, kiip, kic \\
\textbf{ci} & /ch/            & Affricate (as in church): ciald \\
\textbf{u}  & /w/             & Semivowel substitution for \textit{w}: uoter \\
\textbf{i}  & /y/             & Semivowel substitution for \textit{y}: ies \\
\textbf{sh} & /sh/            & As in ship, cash \\
\textbf{zh} & /zh/            & As in vision: vizhun \\
\bottomrule
\end{tabularx}
\caption{Non-standard Consonant Mappings}
\end{table}

\textbf{Aspiration (\me{k} vs \me{c}):} The aspirated /k\textsuperscript{h}/ uses \me{k}; unaspirated uses \me{c}. Following sibilants, \me{c} is mandated (\me{scan}, \me{scip}).

\subsection{Orthographic Normalization}
Historical orthographic artifacts are removed by rule: silent vowels are nullified, \me{-tion} becomes \me{-shun}, \me{ck} resolves by aspiration, \me{x} decomposes to \me{cs}, and double consonants are collapsed.

%---
\section{Morphology and Syntax}

\textbf{Core Constraint:} Lexical items never undergo morphological inflection. Grammatical relationships are managed exclusively through operators.

\subsection{Nouns and Quantitative Prefixes}
Morphological plurals are abolished. Quantity is mandated explicitly.

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{lXl}
\toprule
\textbf{Prefix} & \textbf{Semantic Value} & \textbf{Example} \\
\midrule
\me{1}  & Singular / Unitary  & \me{1cat} \\
\me{3}  & Exact enumeration   & \me{3cat} \\
\me{*}  & Universal / Generic & \me{*cat} \\
\me{\~{}5} & Approximate    & \me{\~{}5cat} \\
\me{0}  & Null set            & \me{0cat} \\
\bottomrule
\end{tabularx}
\caption{Quantitative Prefix System}
\end{table}

\textbf{The Literalization Operator (\me{\_}):} Hyphens (\me{-}) denote culturally unified compounds (\me{haus-fuud} = restaurant). Underscores (\me{\_}) force compositional literal meaning (\me{haus\_fuud} = a house built from food). This arrests lexicalization and semantic drift.

\textbf{Proper Noun Encapsulation:} Proper nouns retain canonical native orthography. Initial capitalization acts as an escape character suspending local phonetic rules: \me{1person Isabella}, \me{1cuntri France}.

\textbf{Fractions and Decimals:} The \me{/} (slais) and \me{.} (point) operators standardize spoken mathematics: \me{1/2apel} (half an apple), \me{0.5} (zero point five).

\subsection{Verbs and Temporal Mechanics}

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Operator} & \textbf{Temporal Value} \\
\midrule
\textit{(null)} & Present continuous (\me{iit}) \\
\me{1} & Present discrete (\me{i1iit}) \\
\me{*} & Habitual / Universal (\me{*iit}) \\
\me{-0s} & Immediate past (\me{-0s sii}) \\
\me{-1d} & Past offset, one day (\me{-1d iit}) \\
\me{:5Y} & Duration, 5 years (\me{:5Y studi}) \\
\me{+1d} & Future offset, one day (\me{+1d gou}) \\
\me{2026-12-25} & Absolute timestamp \\
\bottomrule
\end{tabularx}
\caption{Temporal Prefix System}
\end{table}

\textbf{Temporal Stacking:} Left-to-right operator chaining enables algebraic timeline construction: \me{i-0s -1d iit} = ``I had already eaten yesterday.''

\subsection{Scalar Intensity}
Adjectives post-modify nouns. Intensity is modified via combinators: \me{>big} (very big), \me{>>big} (extremely big), \me{<hot} (slightly hot).

\subsection{Pronominal Indexing}
Single-character spatial indices: \me{i} (I), \me{u} (you), \me{h} (he), \me{s} (she), \me{t} (it). Plurality inherits the quantitative prefix system (\me{5i} = we five). The pronominal index agglutinates directly to the temporal operator: \me{h1laic}, \me{u:2h uorc}.

\subsection{Negation}
Negation is expressed by the invariant prefix \me{no} prepended to the verb:
\me{i1no laic 1cat} (I do not like the cat).

\subsection{Clause Anchors (\me{co} / \me{oc})}
When nesting two or more clauses, \me{co} (clause open) and \me{oc} (clause close) act as explicit stack managers for the human phonological loop, preventing center-embedding failures.

\subsection{Echo-Tagging (Noisy Protocol)}
Per Shannon's Theorem$^{[1]}$, redundancy may be artificially inflated to 100\% by echoing the quantitative prefix as a full word after the root: \me{3boi trii 1uoc nau}.

\subsection{Formal Syntax (Backus-Naur Form)}
The absolute regularity of MinEnglish permits algorithmic validation:

\begin{lstlisting}[language=SQL]
<sentence> ::= [<subj>] <verb_block> [<obj>]
<subj>  ::= <noun_phrase> | <pronoun>
<obj>   ::= <noun_phrase> | <pronoun>
<noun_phrase> ::= <qty> <root> [<adjective>]
<verb_block>  ::= [<temporal>] <root>
<temporal>    ::= <time> | <duration> | <unix>
<clause>      ::= "co" <connect> <sentence> "oc"
\end{lstlisting}

%---
\section{Base Lexicon}

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{lXlX}
\toprule
\textbf{English} & \textbf{MinEnglish} & \textbf{English} & \textbf{MinEnglish} \\
\midrule
be / is & \me{bi}   & have   & \me{hav}   \\
do      & \me{du}   & go     & \me{gou}   \\
come    & \me{cum}  & make   & \me{meic}  \\
think   & \me{tinc} & say    & \me{sei}   \\
want    & \me{uant} & need   & \me{niid}  \\
\bottomrule
\end{tabularx}
\caption{Core Fundamentals}
\end{table}

\subsection{Academic \& Computational Data Lexicon}

Where Latinate terms resist compression, MinEnglish employs semantic compounding from root primitives.

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{lXl}
\toprule
\textbf{Domain} & \textbf{English} & \textbf{MinEnglish} \\
\midrule
\textit{Science}    & Hypothesis  & \me{aidia-test}  \\
                    & Algorithm   & \me{rul-mat}     \\
                    & Cell (Bio)  & \me{bocs-laif}   \\
                    & Evolution   & \me{ceinj-taim}  \\
                    & Gravity     & \me{pul-erd}     \\
\midrule
\textit{Computing}  & Database    & \me{daatabeis}   \\
                    & Variable    & \me{ting-ceinj}  \\
                    & System      & \me{net-ting}    \\
                    & Memory      & \me{haus-daata}  \\
                    & Network     & \me{net-lain}    \\
\midrule
\textit{Economics}  & Inflation   & \me{rap-mun}     \\
                    & Contract    & \me{rul-peper}   \\
\midrule
\textit{Medicine}   & Antibiotic  & \me{bac-med}     \\
                    & Diagnosis   & \me{faind-sic}   \\
                    & Surgery     & \me{cut-laif}    \\
\bottomrule
\end{tabularx}
\caption{Academic Domain Lexicon (Representative Sample)}
\end{table}

%---
\section{Comparative Corpus}

The following 35 examples demonstrate MinEnglish performance across diverse linguistic domains. For each example, a back-translation reconstructs compositional root logic. Compression is measured as character count reduction relative to standard English.

\subsection{Simple Statements \& Universal Truths}

\textbf{Ex. 1 --- Simple Statement (45.7\%):} The three-part English determiner + progressive construction collapses to a single quantitative prefix and present-tense default.\\
\me{3cat sit on 1mat big red.} \textit{(25 chars vs. 46)}

\textbf{Ex. 2 --- General Truth (8.7\%):} The universal quantifier \me{*} and generic habitual verb removes the need for ``is'' and ``that''. Low compression due to Latinate nouns.\\
\me{*dog bi *animal loial dat *luv *h's ouner.} \textit{(42 chars vs. 46)}

\subsection{Past Events, Times, and Futures}

\textbf{Ex. 3 --- Past Event (47.5\%):} The temporal vector \me{-1d} replaces ``Yesterday I...'' in two characters. Number-nouns \me{2buc}, \me{3apel} require no articles.\\
\me{i-1d bai 2buc an 3apel @ 1stor.} \textit{(31 chars vs. 59)}

\textbf{Ex. 4 --- Future + Question (48.9\%):} The \me{+1d} prefix replaces the future tense auxiliary ``will''. The bare \me{?} appended at end marks the entire sentence as interrogative.\\
\me{u+1d iit 1diner uid *i?} \textit{(23 chars vs. 45)}

\textbf{Ex. 5 --- Approximation (58.6\%):} The \me{\~{}} operator replaces the multi-word phrase ``About fifty'' with two characters. Absolute 24-hour time avoids ``three o'clock in the afternoon.''\\
\me{\~{}50person ariv @ \~{}15:00.} \textit{(24 chars vs. 58)}

\textbf{Ex. 6 --- Specific DateTime (63.4\%):} The highest compression example. The ISO 8601 timestamp \me{2026-01-15 14:30} replaces 59 English characters describing dates and times conversationally.\\
\me{1miiting 2026-01-15 14:30 bi held.} \textit{(34 chars vs. 93)}

\textbf{Ex. 7 --- Duration + Contrast (38.5\%):} The \me{:5Y} duration scalar replaces the English perfect ``have been ... for five years''. The modal \me{can} precedes the verb directly.\\
\me{i:5Y studi inglish but i no can spiic fluuent.} \textit{(48 chars vs. 78)}

\subsection{Complex Modifiers \& Voice}

\textbf{Ex. 8 --- Intensity \& Compound Words (47.1\%):} Double-combinator \me{>>niu} expresses ``extremely new''. Slash compound \me{fuud/haus} replaces ``restaurant'' (10 chars) with 8.\\
\me{*i+1d gou tu 1fuud/haus >>niu @ 19:00.} \textit{(38 chars vs. 78)}

\textbf{Ex. 9 --- Passive Voice (47.5\%):} Emphatic double-exclamation \me{!!uac} collapses ``Look out!'' The passive is formed by fronting the object: \me{1uindou...bai 1man}.\\
\me{!!uac! 1uindou glas -0s breic bai 1man angri.} \textit{(31 chars vs. 60)}

\textbf{Ex. 10 --- Modality \& Possessive (29.2\%):} The modal \me{must} and \me{can} attach directly to the pronominal temporal operator. Genitive \me{s'} for feminine avoids double sibilant.\\
\me{s1must giv s' buc tu h bifor s1can start 1uorc.} \textit{(50 chars vs. 61)}

\subsection{Conditionals, Logic \& Commands}

\textbf{Ex. 11 --- Complex Conditional (31.1\%):} The conditional \me{if} directly precedes the clause. Negation \me{no} and the object marker \me{tu} prevent syntactic ambiguity.\\
\me{if u1no stop tu run, u1fol an breic tu u's 2leg.} \textit{(49 chars vs. 61)}

\textbf{Ex. 12 --- Conjunction Reduction (35.2\%):} Conjunction Reduction drops the subject after \me{an}. The habitual \me{*iit} replaces ``eat one piece of toast''.\\
\me{*morning i*drinc 2cup cofi an *iit 1piis toust bifor uorc.} \textit{(56 chars vs. 82)}

\textbf{Ex. 13 --- Instructions (42.9\%):} Recipe-style commands use bare verbs. The universal \me{*ting} replaces ``everything'' in 5 characters.\\
\me{teic 3eg, ad 2cup flaur, an mics *ting for 10m.} \textit{(48 chars vs. 84)}

\textbf{Ex. 14 --- Polite Request (32.7\%):} The deference operator \me{\~{}} prepended to the pronoun encodes politeness without auxiliary verbs: \me{\~{}u1giv} = ``Could you please give''.\\
\me{\~{}u1giv i 2glas uoter, i bi >tersti.} \textit{(35 chars vs. 55)}

\subsection{Dialogue \& Description}

\textbf{Ex. 15 --- Conversation (34.4\%):} The interrogative prefix \me{haum} replaces ``how many''. Direct number-nouns \me{2boi}, \me{1gerl} need no articles or pluralization morphemes.\\
\me{"haum ciald u hav?" "i hav 2boi an 1gerl."} \textit{(42 chars vs. 64)}

\textbf{Ex. 16 --- Narrative (30.6\%):} Temporal vector \me{-1Y} handles ``Last year''. The \me{cos} replaces ``because she wanted to understand''. Compression limited by Latinate nouns.\\
\me{s-1Y travel tu 5cuntri diferent an lern 3languij niu cos s uant understand 1uerld mor gud.} \textit{(93 chars vs. 134)}

\textbf{Ex. 17 --- Emotions \& Probability (41.0\%):} The combined modal \me{\~{}can} expresses probabilistic future action. Focus operator \me{!} before \me{h} specifies that the negation targets the subject.\\
\me{i+1d \~{}can gou tu 1parti, but s1luv h an !h 1no luv s.} \textit{(55 chars vs. 69)}

\textbf{Ex. 18 --- Family Description (8.5\%):} Lowest compression example. The Latinate words \me{hospital} and \me{iuniversiti} maintain protracted phonetic transcriptions that resist operator optimization.\\
\me{i's bruder *uorc @ 1hospital an i's sister *studi @ 1iuniversiti.} \textit{(65 chars vs. 71)}

\textbf{Ex. 19 --- Weather (27.1\%):} Prepositional verb \me{1in 1parc} avoids ``be in''. Temporal stacking \me{-0s start} encodes ``when it started'' in four characters.\\
\me{*ciald -0s 1in 1parc uen t-0s start rein >hevi.} \textit{(49 chars vs. 70)}

\textbf{Ex. 20 --- Price Comparison (21.4\%):} Interrogative marker \me{?haum} fronted on the sentence. Double combinator \me{>>ecspensiv} replaces ``too expensive''.\\
\me{?haum dis car cost? t bi >>ecspensiv for i.} \textit{(44 chars vs. 56)}

\subsection{Advanced Domains}

\textbf{Ex. 21 --- Science Fiction (35.6\%):} Slash compound \me{ship-speis} (spaceship). Immediate past \me{-0s} encodes ``launched''. Hyphen compound \me{caunt-daun} (countdown).\\
\me{1ship-speis -0s lonci in 1orbit after 1caunt-daun hic 0.} \textit{(43 chars vs. 67)}

\textbf{Ex. 22 --- Philosophy (25.0\%):} The bare infinitive \me{bi} replaces the English ``to be''. The future-time token \me{+taim} generalizes ``eventually''.\\
\me{bi or no bi, dat bi 1cuesciun *i must anser in +taim.} \textit{(53 chars vs. 71)}

\textbf{Ex. 23 --- Business Strategy (28.7\%):} Possessive \me{*i's compani} (our company). Compound \me{cein-suplai} (supply chain). Future vector \me{+1Y} replaces ``next year''.\\
\me{*i's compani niid optimaiz t's cein-suplai tu macsimaiz *marjin-profit +1Y.} \textit{(59 chars vs. 83)}

\textbf{Ex. 24 --- Politics (29.5\%):} Immediate past \me{-0s} reduces ``announced'' to two chars. Relative clause \me{dat +1d apect} avoids the ``that will affect'' construction.\\
\me{1prezident -0s anauns 1nolisi-tacs niu dat +1d apect *pamili clas-midl.} \textit{(55 chars vs. 78)}

\textbf{Ex. 25 --- Medicine (36.9\%):} Compound-derived root \me{bac-med} (antibiotics). Compound \me{bac-sic} (bacterial infection). Intensity \me{>siirius} (severe).\\
\me{1doctor -0s prescraib tu *bac-med tu triit 1bac-sic >siirius 1peishent's.} \textit{(53 chars vs. 84)}

\textbf{Ex. 26 --- Real Estate (34.2\%):} Compound \me{ruum-bed} (bedroom). Compound \me{iard-bac} (backyard). Present context marker \me{nau} replaces ``currently''.\\
\me{1haus 3ruum-bed uid 1iard-bac big 1in 1marcet for rent nau.} \textit{(54 chars vs. 82)}

\textbf{Ex. 27 --- Programming (32.9\%):} Compound \me{sistem-bacup} (backup system). The operator \me{otoristor} is a root-compression of ``automatically restore''.\\
\me{if 1server craish, 1sistem-bacup +1d otoristor 1daatabeis.} \textit{(54 chars vs. 81)}

\textbf{Ex. 28 --- Environment (34.0\%):} Inverted compound \me{iurm-gloubal} (global warming) reorders to verb-first. Comparative \me{>fast dan} replaces ``faster than''.\\
\me{iurm-gloubal 1melt *cap-ais poular >fast dan *saientist -1d pridiict.} \textit{(51 chars vs. 78)}

\textbf{Ex. 29 --- Literature / Metaphor (29.2\%):} The immediate past \me{-0s} applies to both verbs (\me{uisper}, \me{raiz}) simultaneously. Adjective post-modifier \me{silver} needs no comma.\\
\me{1uind -0s uisper tru *trii >ould az 1muun silver -0s raiz in 1scai.} \textit{(56 chars vs. 80)}

\textbf{Ex. 30 --- Logistics (43.4\%):} The compound \me{truc-diliveri} (delivery truck). The preposition \me{btuiin} handles exact time ranges without ``o'clock''.\\
\me{1truc-diliveri +1d ariv uid u's pacij btuiin 12:00 an 14:00.} \textit{(44 chars vs. 78)}

\textbf{Ex. 31 --- Legal (23.3\%):} Domain-specific transliterations (\me{diipendant}, \me{carj}) resist operator optimization. Second-lowest compression in the corpus.\\
\me{1diipendant -0s formal pliid no gilti tu *carj prizent bai 1cort.} \textit{(64 chars vs. 80)}

\textbf{Ex. 32 --- Cooking (34.3\%):} Bare imperative commands. Object marker \me{tu} required before \me{1uoter} to prevent ambiguity with the directive. Generic \me{*ster} (stir continuously).\\
\me{boil tu 1uoter, ad tu 1pinci solt, an *ster until 1sos >tic.} \textit{(60 chars vs. 82)}

\textbf{Ex. 33 --- Emotional Confession (40.5\%):} Temporal stacking \me{-0s -1d} encodes ``have never before''. Universal negation \me{*no fiil} handles ``never felt''.\\
\me{i-0s -1d *no fiil dis ue abaut *person bifor, u -1d ceinj i's laif >hol.} \textit{(70 chars vs. 75)}

\textbf{Ex. 34 --- Navigation (38.8\%):} Ordinal \me{tuu} in place of ``second'' (traffic light). The interrogative relative \me{u1sii} resolves the trailing ``you see'' reference.\\
\me{gou streit for 2mail, den tern left @ 1lait-trapic tuu u1sii.} \textit{(46 chars vs. 78)}

\textbf{Ex. 35 --- Abstract Economics (33.3\%):} The compound \me{rap-mun} collapses ``hyperinflation'' (14 chars) to 7. Double combinator \me{>>} expresses ``hyper-'' intensity.\\
\me{>>rap-mun 1distroi tu pauer-bai an >siirius holt tu divelopment-mun.} \textit{(54 chars vs. 81)}

%---
\section{Comparative Information Density Analysis}

Transmission efficiency was modeled against the 35-sentence corpus above. Efficiency is measured as character count reduction.

\begin{table}[h!]
\centering\small
\begin{tabularx}{\linewidth}{lrX}
\toprule
\textbf{Example} & \textbf{Reduction} & \textbf{Key Mechanism} \\
\midrule
Ex. 6: Timestamp    & 63.4\% & ISO 8601 collapses 59 chars \\
Ex. 5: Approximation & 58.6\% & \me{\~{}50person} vs. ``About fifty people'' \\
Ex. 4: Future Q     & 48.9\% & \me{+1d} eliminates ``will'' chain \\
\midrule
Ex. 18: Family      & 8.5\%  & Latinate nouns resist compression \\
Ex. 31: Legal       & 23.3\% & Domain transliterations are long \\
\bottomrule
\end{tabularx}
\caption{Top and Bottom Compression Vectors}
\end{table}

\textbf{Corpus Density Metric (n=35):}
\begin{itemize}[leftmargin=*]
    \item Total English characters: \textbf{2633}
    \item Total MinEnglish characters: \textbf{1711}
    \item \textbf{Average reduction: 35.0\%}
\end{itemize}

%---
\section{Theoretical Limitations: The Human Biological Factor}

The structural additions of v2.0 (Literalization \me{\_}, Clause Anchors \me{co}/\me{oc}, Echo-Tagging) address standard linguistic inefficiencies. However, the primary limitation of MinEnglish is not structural but biological.

Mammalian communicative pathways evolved to facilitate social cohesion, hierarchical navigation, and emotional subtext through productive ambiguity$^{[2]}$. MinEnglish enforces an objective, algebraic paradigm. The absolute requirement for temporal precision, numerical exactness, and the prohibition of metaphorical obfuscation inherently conflicts with natural human cognitive behavior. Widespread adoption would likely prompt measurable psychological resistance, as the protocol disables the linguistic mechanisms humans rely on to manage sociological complexity.

%---
\section*{References}
\begin{enumerate}
\item Shannon, C. E. (1948). A mathematical theory of communication. \textit{Bell System Technical Journal}, 27(3), 379--423.
\item Miller, G. A. (1956). The magical number seven, plus or minus two: Some limits on our capacity for processing information. \textit{Psychological Review}, 63(2), 81--97.
\item Chomsky, N. (1957). \textit{Syntactic Structures}. Mouton.
\item Zipf, G. K. (1949). \textit{Human Behavior and the Principle of Least Effort}. Addison-Wesley.
\end{enumerate}

%---
\begin{appendices}

\section*{MIT License}

\begin{lstlisting}
Copyright (c) 2026 Dan Micsa, PhD

Permission is hereby granted, free of charge,
to any person obtaining a copy of this language
specification and associated documentation files
(the "Specification"), to deal in the Specification
without restriction, including without limitation
the rights to use, copy, modify, merge, publish,
distribute, sublicense, and/or sell copies of the
Specification, and to permit persons to whom the
Specification is furnished to do so, subject to
the following conditions:

The above copyright notice and this permission
notice shall be included in all copies or
substantial portions of the Specification.

THE SPECIFICATION IS PROVIDED "AS IS", WITHOUT
WARRANTY OF ANY KIND. IN NO EVENT SHALL THE AUTHORS
BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY HOWEVER CAUSED.
\end{lstlisting}

\end{appendices}

\end{document}
